Symbolic linearization of equations of motion of constrained multibody systems

### 2013-07-04 PetersonGedeHubbard2012.tex
Incorporated Mont's changes.

\begin{abstract}
% Context
Many common mechanical systems have configuration or velocity constraints.
% Need  (what we have and what we want)
Analyses of such systems often require linearized forms of the motion
equations.
% Task
To address this issue, we developed a procedure for organizing the constraint
and motion equations and their subsequent linearization.
This procedure was specifically developed for equations of motion generated by
Kane's method, but is compatible with any method as long as the following are
true: only ordinary differential equations are required to describe the system,
and independent and dependent states remain in the final equations (i.e. no
DAEs, no substitution of variables).
% Object of the document
Following a brief review of Kane's method and the structure of the equations it
generates, we present the procedure to symbolically linearize the nonlinear
motion equations of constrained multibody systems, and illustrate it with the
% TODO: change name of example
familiar example of the rolling disc.

@@

\section{Introduction}
\label{sec:intro}
Designing a feedback controller or performing stability analysis are two common
analyses performed on models of multibody systems.
A linear form of the equations of motion is required to apply the most
frequently used techniques for the analyses to such systems.
Normally, obtaining the linearized equations of motion is as simple as taking
the Jacobian of the right hand side function of the equations of motion with
respect to the states.
However, when the multibody system in questions has constraints, this process
is more complicated.
If one has used Kane's method \cite{Kane1985} to generate the equations for a
constrained system, the resulting equations will have a particular form: a set
of ordinary differential equations containing both the independent and
dependent states.
Simple inspection of the equations does not reveal that they represent a
constrained system - their form is similar to the equations of an unconstrained
system.
It is critical to understand linearizing these equations by computing the
Jacobian gives incorrect results.
This paper provides a way to correctly generate the linear equations of motion
for constrained systems which have been found using Kane's method (or any other
method which results in equations with the same form).

@@

% TODO: Improve literature review.
Bottema appears to be the first author to show that special considerations are
needed for linearizing nonholonomic systems \cite{Bottema1949}.
While other authors have examined linearization of nonholonomic systems, we
have found isssues with applying any of them to equations generated by Kane's
method.
Kang et al. \cite{Kang2003} and Negrut and Ortiz \cite{Negrut2006} have both
explored linearization, but these methods have been developed for equations of
motion found using Lagrange's method, which contain Lagrange multipliers (which
are not present in equations of motion generated from Kane's method).
Minaker and Rieveley \cite{Minaker2010} and Schwab and Meijaard
\cite{Schwab2003} both have developed techniques for generating equations of
motion (and linear forms thereof), however both techniques are more numerical
in nature - they do not provide symbolic equations of motion or symbolic
linearized equations.
These processes, along with others such as Neuman's %TODO: fix reference
, formulate the linear equations as a step in the equations of motion
generation process.
They are not capable of linearizing existing equations of motion.

@@

% TODO:rewrite organization of paper
The use of Kane's method in generating equations of motion is reviewed first,
in order to properly orient the reader to the format and some of the qualities
of the generated equations.
Kane's method is then used to generate the equations of motion for a simple
system, 
%TODO: update example name 
a rolling disk with a trailing arm.
The need for a different linearization technique is also demonstrated with this
model.
In response to this need, we present the newly developed linearization
procedure and its derivation.
This new linearization procedure is then performed on the example system to
provide an example of its application.
Finally, we discuss other details and nuances of the procedure and its use.

@@

When using Lagrange's method, generalized coordinates are used to describe the
configuration of a system while the time derivatives of the gen. coordinates
describe the velocities of a system.
Kane's method allows for the velocities of a system to be written in terms of
generalized speeds, which are not required to be the time derivatives of
coordinates (although such a definition is permitted).

The benefit of using both generalized speeds and coordinates can be seen in
the following example.
If the orientation of a rigid body were to be described using using Euler
parameters (quaternions), its angular velocity is relatively complicated when
using the time derivatives of the coordinates.
Using generalized speeds allows for the angular velocity to be written in the
much simpler form
\begin{align}
\label{eq:ex1}
{^N}\bar{\omega}^B = \it{\omega}_1 \hat{b}_1 + \it{\omega}_2 \hat{b}_2 +
\it{\omega}_3 \hat{b}_3
\end{align}
or, the angular velocity of rigid body $B$ in the reference frame $N$ is the
defined as the sum of three generalized-speed/basis-vector products.
The derivatives of these generalized speeds will then appear in the equations
of motion, rather than the twice time differentiated generalized coordinates
($\ddot{\mathbf{q}}$'s).

A downside of allowing such definitions is clear when formulating the kinematic
differential equations, which relate the rate of change of the generalized
coordinates to the generalized speeds.
These equations become much more complicated, but this is usually offset by the
significantly simpler equations of motion.

Use and understanding of Kane's method requires another concept, partial
velocities.
Partial velocities are defined as the partial derivatives of the velocity
vectors with respect to the generalized speeds.
They are defined for both translational and rotational velocities.
Using the angular velocity of body $B$ in reference frame $N$ as shown in
(\ref{eq:ex1}), the partial velocities can be written
\[
{^N}\bar{\omega}^B_{\omega_1} = \hat{b}_1 \quad \quad
{^N}\bar{\omega}^B_{\omega_2} = \hat{b}_2 \quad \quad
{^N}\bar{\omega}^B_{\omega_3} = \hat{b}_3
\]
and it can be seen that the partial velocities represent the direction of
motion associated with each generalized speed.

By taking the dot product of each partial velocity and both sides of Newton's
second law, $\bar{F}=m\bar{a}$ (or the rotational equivalent), only terms that
are parallel to the partial velocities are left in the equation.
One equation is generated for each generalized speed.

An important byproduct of this is that internal constraint forces
(non-contributing forces) imposed by constructs such as pin or sliding joints
no longer need to be considered.

In summary, Kane's method allows for velocities to be defined in terms of
generalized speeds, allowing for simpler velocity expressions.
The dot product of each partial velocity (the direction of motion associated
with each generalized speed) with Newton's second law (or Euler's equations)
removes all terms which are not related to each generalized speed.
This generally results in simple equations of motion, which are easier to form.

@@

In order to properly apply Kane's method, the velocities of each component in
the system need to be written in the following form.

@@

In these equations, each term $\bar{v}_t$ or $\bar{\omega}_t$ represents the
velocity or angular velocity component which is a prescribed function of time,
and has no dependence on generalized speeds (e.g., a crank
driven at a fixed rate).

For this system, the configuration constraints are defined as:
\begin{align}
\label{eq:configuration_constraints}
\mathbf{f}_c(\mathbf{q}, t) = \mathbf{0}
\end{align}
where the terms $\mathbf{f}_c$ and $\mathbf{0}$ have the shape $l \times 1$.
The velocity constraints are defined as:
\begin{align}
\label{eq:velocity_constraints}
\mathbf{f}_v(\mathbf{q}, \mathbf{u}, t) = \mathbf{0}
\end{align}
where the terms $\mathbf{f}_v$ and $\mathbf{0}$ have the shape $m \times 1$.
It is important that the velocity constraints include the time-differentiated
configuration constraints or equivalent constraints which produce the same
behavior.
Furthermore, velocity constraints can not be written with $\dot{\mathbf{q}}$
terms (the time derivatives of the coordinates).
Acceleration constraints need to be defined as well.
Typically, they will simply by time-differentiated velocity constraints and
have the form:
\begin{align}
\label{eq:acceleration_constraints}
\mathbf{f}_a(\mathbf{q}, \dot{\mathbf{q}}, \mathbf{u}, \dot{\mathbf{u}}, t) =
\mathbf{0}
\end{align}

@@

To simplify notation, it is assumed that the generalized speeds are already
ordered in a favorable manner.
\begin{align}
\label{eq:speeds_independent_dependent}
\mathbf{u} =
\begin{bmatrix}
\mathbf{u}_i \\
\mathbf{u}_d
\end{bmatrix}
\end{align}
The matrix $\mathbf{B}$ from (\ref{eq:constraint_B}) also needs to be
partitioned.
\begin{align}
\label{eq:constraint_B_independent_dependent}
\mathbf{0} &= \mathbf{B} \mathbf{u}\\
\mathbf{0} &= \begin{bmatrix}
\mathbf{B}_i & \mathbf{B}_d
\end{bmatrix} \begin{bmatrix}
\mathbf{u}_i \\ \mathbf{u}_d
\end{bmatrix}
\end{align}
where $\mathbf{B}_i$ has the shape $m \times o - m$ and $\mathbf{B}_d$ has the
shape $m \times m$.
This equation can then be rewritten as
\begin{align}
\label{eq:constraint_A}
\mathbf{B}_i \mathbf{u}_i + \mathbf{B}_d \mathbf{u}_d = \mathbf{0} \\
\mathbf{B}_d \mathbf{u}_d = - \mathbf{B}_i \mathbf{u}_i \\
\mathbf{u}_d = -\mathbf{B}_d^{-1} \mathbf{B}_i \mathbf{u}_i \\
\mathbf{A} \triangleq - \mathbf{B}_d^{-1} \mathbf{B}_i \\
\mathbf{u}_d = \mathbf{A} \mathbf{u}_i
\end{align}
where $\mathbf{A}$ has the shape $m \times m$.

@@

After constraining the equations of motion, we have reduced the number of
equations from $o$ to $o-m$; but the equations still contain $o$ unknown terms
that need to be solved for ($\dot{\mathbf{u}}_1,...,\dot{\mathbf{u}}_o$).
In order to solve these second order differential equations for the
$\dot{\mathbf{u}}$ terms, additional equations are needed.
The acceleration constrain equations fill this role, as there are $m$ of these
equations, giving:
\begin{align}
\label{eq:F_Fstar_f_a}
\begin{bmatrix}
\tilde{\mathbf{F}} + \tilde{\mathbf{F}}^* \\
\mathbf{f}_a (\mathbf{q}, \dot{\mathbf{q}}, \mathbf{u}, \dot{\mathbf{u}}, t)
\end{bmatrix} = \mathbf{0}
\end{align}
which gives a total of $o$ equations, enough to be able to solve for the
$\dot{\mathbf{u}}$'s.

>>>>

\begin{abstract}
% Context
Many common dynamic mechanical systems have configuration or velocity constraints.
% Need  (what we have and what we want)
Analyses of such systems often require linearized forms of the motion
equations.
% Task
To address this issue, we developed a procedure for organizing the constraint
and motion equations and their subsequent linearization.
This procedure was specifically developed for equations of motion generated by
Kane's method, but is compatible with any method as long as the following are
true: only ordinary differential equations are required to describe the system,
and independent and dependent states remain in the final equations (i.e. no
DAEs, no substitution of variables).
% Object of the document
Following a brief review of Kane's method and the structure of the equations it
generates, we present the procedure to symbolically linearize the nonlinear
motion equations of constrained multibody systems, and illustrate it with the
% TODO: change name of example
familiar example of the rolling disc.

@@

\section{Introduction}
\label{sec:intro}
Designing a feedback controller or performing stability analysis are two common
analyses performed on models of multibody systems.
A linear form of the equations of motion is required to apply the most
frequently used techniques for the analyses to such systems.
Normally, obtaining the linearized equations of motion is as simple as
calculating the Jacobian of the right hand side function of the equations of
motion with respect to the states.
However, when the multibody system in questions has constraints, this process
is more complicated.
If one has used Kane's method \cite{Kane1985} to generate the equations for a
constrained system, the resulting equations will have a particular form: a set
of ordinary differential equations containing both the independent and
dependent states.
Simple inspection of the equations does not reveal that they represent a
constrained system - their form is similar to the equations of an unconstrained
system.
It is critical to understand linearizing these equations by computing the
Jacobian gives incorrect results.
This paper provides a way to correctly generate the linear equations of motion
for constrained systems which have been found using Kane's method (or any other
method which results in equations with the same form).

@@

% TODO: Improve literature review.
Bottema appears to be the first author to have shown that special
considerations are needed for linearizing nonholonomic systems
\cite{Bottema1949}.
While other authors have examined linearization of nonholonomic systems, we
have found isssues with applying any of them to equations generated by Kane's
method.
Kang et al. \cite{Kang2003} and Negrut and Ortiz \cite{Negrut2006} have both
explored linearization, but these methods have been developed for equations of
motion found using Lagrange's method, which contain Lagrange multipliers (which
are not present in equations of motion generated from Kane's method).
Minaker and Rieveley \cite{Minaker2010} and Schwab and Meijaard
\cite{Schwab2003} both have developed techniques for generating equations of
motion (and linear forms thereof), however both techniques are more numerical
in nature - they do not provide symbolic equations of motion or symbolic
linearized equations.
These processes, along with others such as Neuman's %TODO: fix reference
, formulate the linear equations as a step in the equations of motion
generation process.
They are not capable of linearizing existing equations of motion.

@@

% TODO:rewrite organization of paper
The use of Kane's method in generating equations of motion is reviewed first,
in order to properly orient the reader to the format and some of the qualities
of the generated equations.
Kane's method is then used to generate the equations of motion for a simple
system,
%TODO: update example name
a rolling disk with a trailing arm.
The need for a different linearization technique is also demonstrated with this
example.
In response to this need, we present the newly developed linearization
procedure and its derivation.
This new linearization procedure is then performed on the system to provide an
example of its application.
Finally, we discuss other details and nuances of the procedure and its use.

@@

When using Lagrange's method, generalized coordinates are used to describe the
configuration of a system while the time derivatives of the generalized
coordinates describe the velocities of a system.
Kane's method allows for the velocities of a system to be written in terms of
generalized speeds, which are not required to be the time derivatives of
coordinates (although such a definition is permitted).

The benefit of choosing generalized speeds to be other than simply the
derivatives of the corresponding generalized coordinates can be seen in the
following example.
If the orientation of a rigid body were to be described using using Euler
parameters as coordinates, its angular velocity is relatively complicated when
using the time derivatives of the coordinates and requires a velocity
constraint.
Using generalized speeds allows for the angular velocity to be written in the
much simpler form
\begin{align}
\label{eq:ex1}
{^N}\bar{\omega}^B = \it{\omega}_1 \hat{b}_1 + \it{\omega}_2 \hat{b}_2 +
\it{\omega}_3 \hat{b}_3
\end{align}
or, the angular velocity of rigid body $B$ in the reference frame $N$ is the
defined as the sum of three generalized-speed/basis-vector products.
The derivatives of these generalized speeds will then appear in the equations
of motion, rather than the twice time differentiated generalized coordinates
($\ddot{\mathbf{q}}$'s).

A drawback of allowing such definitions is clear when formulating the kinematic
differential equations, which relate the rate of change of the generalized
coordinates to the generalized speeds.
These equations become much more complicated, but this is usually offset by the
significantly simpler equations of motion.

Use and understanding of Kane's method requires another concept, partial
velocities.
Partial velocities are defined as the partial derivatives of the velocity
vectors with respect to the generalized speeds and are defined for both
translational and rotational velocities.
Using the angular velocity of body $B$ in reference frame $N$ as shown in
(\ref{eq:ex1}), the three partial velocities can be written
\[
{^N}\bar{\omega}^B_{\omega_1} = \hat{b}_1 \quad \quad
{^N}\bar{\omega}^B_{\omega_2} = \hat{b}_2 \quad \quad
{^N}\bar{\omega}^B_{\omega_3} = \hat{b}_3
\]
and it can be seen that the partial velocities represent the direction of
motion associated with each generalized speed.

By taking the dot product of each partial velocity and both sides of Newton's
second law, $\bar{F}=m\bar{a}$ (or the rotational equivalent), only terms that
are parallel to the partial velocities remain in the equation.
One equation is generated for each generalized speed.

An important byproduct of this choice is that internal constraint forces
(non-contributing forces) imposed by objects such as pin or sliding joints
no longer need to be considered.

In summary, Kane's method allows for velocities to be defined in terms of
generalized speeds, allowing for simpler velocity expressions.
The dot product of each partial velocity (the direction of motion associated
with each generalized speed) with Newton's second law (or Euler's equations)
removes all terms which are not related to the rate of change of each
generalized speed.
This generally results in simple equations of motion, which are easier to form.

@@

In order to properly apply Kane's method, the velocities of each component in
the system need to be written as linear combinations of the generalized speeds.

@@

In these equations, each term $\bar{v}_t$ or $\bar{\omega}_t$ represents a
velocity or angular velocity component which is a prescribed function of time,
and has no dependence on generalized speeds (e.g., a crank
driven at a specified rate).

For this system, the configuration constraints are defined as:
\begin{align}
\label{eq:configuration_constraints}
\mathbf{f}_c(\mathbf{q}, t) = \mathbf{0}
\end{align}
where the terms $\mathbf{f}_c$ and $\mathbf{0}$ have the shape $l \times 1$.
The velocity constraints are defined as:
\begin{align}
\label{eq:velocity_constraints}
\mathbf{f}_v(\mathbf{q}, \mathbf{u}, t) = \mathbf{0}
\end{align}
where the terms $\mathbf{f}_v$ and $\mathbf{0}$ have the shape $m \times 1$.
It is important that the velocity constraints include the time-differentiated
configuration constraints or equivalent constraints which produce the same
behavior.
Furthermore, velocity constraints should not be written with $\dot{\mathbf{q}}$
terms (the time derivatives of the coordinates).
Acceleration constraints need to be defined as well.
Typically, they will simply by time-differentiated velocity constraints and
have the form:
\begin{align}
\label{eq:acceleration_constraints}
\mathbf{f}_a(\mathbf{q}, \dot{\mathbf{q}}, \mathbf{u}, \dot{\mathbf{u}}, t) =
\mathbf{0}
\end{align}

@@

To simplify notation, it is assumed that the generalized speeds are already
ordered in a favorable manner.
\begin{align}
\label{eq:speeds_independent_dependent}
\mathbf{u} =
\begin{bmatrix}
\mathbf{u}_i \\
\mathbf{u}_d
\end{bmatrix}
\end{align}
The matrix $\mathbf{B}$ from (\ref{eq:constraint_B}) also needs to be
partitioned.
\begin{align}
\label{eq:constraint_B_independent_dependent}
\mathbf{0} &= \mathbf{B} \mathbf{u}\\
\mathbf{0} &= \begin{bmatrix}
\mathbf{B}_i & \mathbf{B}_d
\end{bmatrix} \begin{bmatrix}
\mathbf{u}_i \\ \mathbf{u}_d
\end{bmatrix}
\end{align}
where $\mathbf{B}_i$ has the shape $m \times o - m$ and $\mathbf{B}_d$ has the
shape $m \times m$.
This equation can then be rewritten as
\begin{align}
\label{eq:constraint_A}
\mathbf{B}_i \mathbf{u}_i + \mathbf{B}_d \mathbf{u}_d &= \mathbf{0} \\
\mathbf{B}_d \mathbf{u}_d &= - \mathbf{B}_i \mathbf{u}_i \\
\mathbf{u}_d &= -\mathbf{B}_d^{-1} \mathbf{B}_i \mathbf{u}_i \\
\mathbf{A} &\triangleq - \mathbf{B}_d^{-1} \mathbf{B}_i \\
\mathbf{u}_d &= \mathbf{A} \mathbf{u}_i
\end{align}
where $\mathbf{A}$ has the shape $m \times m$.

@@

After constraining the equations of motion, we have reduced the number of
equations from $o$ to $o-m$; but the equations still contain $o$ unknown terms
that need to be solved for ($\dot{\mathbf{u}}_1,...,\dot{\mathbf{u}}_o$).
In order to solve these second order differential equations for the
$\dot{\mathbf{u}}$ terms, additional equations are needed.
The acceleration constraint equations fill this role, as there are $m$ of these
equations, giving:
\begin{align}
\label{eq:F_Fstar_f_a}
\begin{bmatrix}
\tilde{\mathbf{F}} + \tilde{\mathbf{F}}^* \\
\mathbf{f}_a (\mathbf{q}, \dot{\mathbf{q}}, \mathbf{u}, \dot{\mathbf{u}}, t)
\end{bmatrix} = \mathbf{0}
\end{align}
which gives a total of $o$ equations, enough to be able to solve for the
$\dot{\mathbf{u}}$'s.
